\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-08 paper.tex
Done up to S2.6

Although we emphasise that further work is required to filter out spuriously inferred recombination events,
we can draw some preliminary conclusions from Figure~\ref{fig:breakpoint-distribution}.
The number of intervals spanning a site (orange curve) is lower at the ends of the genome,
as expected due to the lack of information about recombination with few flanking sites.
In addition, the number of intervals spanning a site often
drops near the beginning of a gene.
This is particularly apparent at the ORF8/N gene interface,
with the N gene containing fewer potential recombination breakpoints
than other genes, in agreement with the results of \cite{Turakhia2022-it}.
Noticeable declines in the number of spanning intervals are also
seen near the beginning of S, M, ORF7a, and ORF8.
These declines are sometimes associated with hypermutated sites (e.g., 27,384 and 28,271
near the beginning of ORF7a and N, respectively), as expected because sites that
undergo mutation at high rates are more likely to differ between the
parents of a recombinant and so provide information about the location of a breakpoint.
This pattern may, however, also be an artefact of sequencing errors causing sites
to appear different between the parents when they are not
(see the discussion of potential errors at site 28,271 below).
In other cases, however, the drop in intervals does not appear to
coincide with hypermutability and may reflect shifts in the actual
rate of recombination between genes.
Indeed, template switching is known to occur at hotspots,
which often involve transcription-regulatory sequences preceding genes
in SARS-CoV-2~\citep{Yang2021-characterizing}
The rate of recombination may also depend on the relative abundance
of different subgenomic RNA intermediates that span different genes,
affecting the availability of templates and influencing the rate of
homologous recombination~\citep{Kim2020-gt,Zou2021-sars}
\sally{[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here).]}

>>>>

Although further work is required to filter
out spuriously inferred recombination events (see
Section~\ref{sec:false_positives})
we can draw some preliminary conclusions from Figure~\ref{fig:breakpoint-distribution}.
The number of intervals spanning a site (orange curve) is lower at the ends of the genome,
as expected due to the lack of information about recombination with few flanking sites.
In addition, the number of intervals spanning a site often
drops near the beginning of a gene.
This is particularly apparent at the ORF8/N gene interface,
with the N gene containing fewer potential recombination breakpoints
than other genes, in agreement with the results of \cite{Turakhia2022-it}.
Noticeable declines in the number of spanning intervals are also
seen near the beginning of S, M, ORF7a, and ORF8.
These declines are sometimes associated with hypermutated sites (e.g., 27,384 and 28,271
near the beginning of ORF7a and N, respectively), as expected because sites that
undergo mutation at high rates are more likely to differ between the
parents of a recombinant and so provide information about the location of a breakpoint.
This pattern may, however, also be an artefact of sequencing errors causing sites
to appear different between the parents when they are not
(see the discussion of potential errors at site 28,271 below).
In other cases, however, the drop in intervals does not appear to
coincide with hypermutability and may reflect shifts in the actual
rate of recombination between genes.
Indeed, template switching is known to occur at hotspots,
which often involve transcription-regulatory sequences preceding genes
in SARS-CoV-2~\citep{Yang2021-characterizing}.
The rate of recombination may also depend on the relative abundance
of different subgenomic RNA intermediates that span different genes,
affecting the availability of templates and influencing the rate of
homologous recombination~\citep{Kim2020-gt,Zou2021-sars}.
\sally{[Sally: KATHERINE, WHAT DO YOU THINK?  These not-full-length subgenomic RNA intermediates may provide a lot of homologous templates for some genes more than other (e.g., fig 3B here).]}
